---
title: "Untitled"
output: html_document
date: "2026-08-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Testing Differences in Thresholds

In most models with ordinal indicators,
the response proportions per level of the ordinal factors is modeled as a "threshold": a value of the standard normal distribution where the probability mass in the left tail corresponds to the response proportion of that level of the ordinal factor.

In these LCA models, however, we instead model response proportions per level of the ordinal factors in terms of threshold differences, which are strictly positive.
This benefits model convergence.
The thresholds are then computed by summing successive differences.

Verify that the parameters are indeed deviances, not thresholds themselves by running `coef(res_final)`.
The function `table_results(res_final)` computes the thresholds by summing the deviances.

We can use a Wald test to test for threshold differences between the classes.
However, to do so, we have to manually specify the sum of parameters.
Thus, to compare the first threshold of the first indicator between the three classes, we could run:

```{r echo = TRUE, eval = FALSE}
res <- readRDS("lca_res.RData")
res_final = res[[3]]
table_results(res[[3]])
wald_test(res_final, "class1.mat_dev[1,1] = class2.mat_dev[1,1] = class3.mat_dev[1,1]")
```

To test if the second threshold of the first variable differs across all three classes, we could use the following more complicated syntax, which adds the deviances:

```{r}
wald_test(res_final, "(class1.mat_dev[1,1]+class1.mat_dev[2,1], class2.mat_dev[1,1]+class2.mat_dev[2,1]) = class3.mat_dev[1,1]+class3.mat_dev[2,1]")
```
