---
title: "Generating syntax for structural equation models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating syntax for structural equation models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidySEM)
library(lavaan)
library(dplyr)
```

`tidySEM` offers a user-friendly, tidy workflow for generating syntax for SEM models. The workflow is top-down, meaning that syntax is generated based on conceptual model elements. In many cases, the generated syntax will suffice - but it is always customizable. The workflow also tries to intelligently guess which variables go together, but these defaults can be overridden.


## The tidySEM workflow

The workflow underlying syntax generation in `tidySEM` is as follows:

1. Give the variables in your `data` object short, informative names, that are easily read by the machine
2. Generate a data dictionary by running `dict <- get_dictionary(data)`
<!--    * The `data_dictionary` called `dict` can be used to generate `keys(dict)`, for use with functions (reliability, factor analysis) from the `psych` library
    * The `data_dictionary` called `dict` can be used to generate syntax for structural equation models in `lavaan` or Mplus (through `MplusAutomation`)-->
3. Convert the `data_dictionary` to a `sem_syntax` object by adding elements of syntax
    * E.g., `dict %>% cfa() -> model`
5. *Optionally*, access the `data_dictionary` and `syntax` elements in the `sem_syntax` object `dictionary(model)` and `syntax(model)`
6. *Optionally*, modify the `data_dictionary` and `syntax` elements in the `sem_syntax` object `dictionary(model) <- ...` and `syntax(model) <- ...`
7. Run an analysis on the generated syntax using `lavaan::sem()` or `MplusAutomation::mplusObject()`

Objects returned by all functions are "tidy" data, i.e., tabular `data.frames`, and can be modified using the familiar suite of functions in the `tidyverse`.

## Example: Running a CFA

### Step 1: Check the variable names

As an example, let's make a graph for a classic `lavaan` tutorial example for CFA. First, we check the data names:

```{r}
df <- HolzingerSwineford1939
names(df)
```

These names are not informative, as the items named `x..` are indicators of three different latent variables. We will rename them accordingly:

```{r}
names(df)[grepl("^x", names(df))] <- c("vis_1", "vis_2", "vis_3", "tex_1", "tex_2", "tex_3", "spe_1", "spe_2", "spe_3")
```

### General rules for naming variables

In general, it is good practice to name variables using the following information:

* Scale name (or if the variable is not part of a scale, observed variable name)
* Measurement occasion (if longitudinal)
* Respondent id (if multiple respondents completed the same scales)
* Scale item number, which `tidySEM` expects to be separated from the remainder of the variable name by a splitting character (e.g., `scale_item`)

Roughly speaking, the elements of the variable name are ordered from "slow-changing" to "fast-changing"; i.e.; there are only a few scales, with possibly several measurement occasions or respondents, and many items.

### Step 2: Generate a dictionary

Thanks to the informative names we've given the data, we can easily construct a data dictionary:

```{r}
dict <- get_dictionary(df)
dict
```

<!--### Optional: Get reliability from psych

Using the data dictionary, we can easily get a keys list for use with functions from the `psych` library, such as `psych::alpha()`:


library(psych)
psych::alpha(df, keys = tidySEM:::keys(dict))

-->

### Step 3: Generate syntax

We can convert the `data_dict` object to `sem_syntax`, simply by adding a first syntax element; in this case, a measurement model:

```{r, eval = FALSE}
dict %>%
  measurement() -> model
model
```

### Optional step 5: Access the dictionary and syntax

The dictionary and syntax can be examined using `dictionary(model)` and `syntax(model)`:


### Optional step 6: Modify the dictionary and syntax

At this stage, we may want to modify the basic syntax slightly. The functions `dictionary(model) <- ...` and `syntax(model) <- ...` can be used to modify the dictionary and syntax:

