---
title: "Tabulating results from structural equation models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tabulating_results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MplusAutomation)
run_mplus <- mplusAvailable(TRUE) == 0
```

```{r setup}
library(tidySEM)
library(lavaan)
library(MplusAutomation)
library(dplyr)
```

`tidySEM` tabulates the results of different types of models in the same uniform way. This facilitates parsing the output into Tables and Figures for publication.

## Output from lavaan

As an example, let's tabulate the results from a classic `lavaan` tutorial example for a multiple group model. First, we run the model:

```{r}
HS.model <- '  visual =~ x1 + x2 + x3
              textual =~ x4 + x5 + x6
              speed   =~ x7 + x8 + x9 '
fit <- cfa(HS.model, 
           data = HolzingerSwineford1939, 
           group = "school")
```
The results can be tabulated using the `lavaan` function `summary()`:

```{r}
summary(fit)
```

Alternatively, we can use the `tidySEM` function `table_results()`:

```{r}
table_results(fit)
```

## Output from Mplus

Now, we'll reproduce the same analysis in Mplus, via the package `MplusAutomation`. This code will only work on your machine if you have Mplus installed and R can find it. First, we run the model:

```{r, eval = FALSE, echo = TRUE}
fit <- mplusModeler(mplusObject(VARIABLE = "grouping IS school (1 = GW 2 = Pas);",
                                MODEL = c("visual BY x1 x2 x3;",
                                          "textual BY x4 x5 x6;",
                                          "speed BY x7 x8 x9;"),
                                usevariables = c(paste0("x", 1:9), "school"),
                                rdata = HolzingerSwineford1939),
                    modelout = "example.inp",
                    run = 1L)
fit$results$parameters
```
```{r, eval = run_mplus, echo = FALSE}
fit <- mplusModeler(mplusObject(VARIABLE = "grouping IS school (1 = GW 2 = Pas);",
                                MODEL = c("visual BY x1 x2 x3;",
                                          "textual BY x4 x5 x6;",
                                          "speed BY x7 x8 x9;"),
                                usevariables = c(paste0("x", 1:9), "school"),
                                rdata = HolzingerSwineford1939),
                    modelout = "example.inp",
                    run = 1L)
file.remove(list.files(pattern = "^example.+(inp|out|dat)$"))
dput(fit$results$parameters, file = "tabres_1.R")

```
```{r eval = TRUE, echo = FALSE}
# Read the results
fit <- list(results = list(parameters = dget("tabres_1.R")))
class(fit) <- c("mplusObject", "list")
class(fit$results) <- c("mplus.model", "list")
fit$results$parameters
```

The results can be tabulated using the `MplusAutomation` function `coef.mplus.model()`:

```{r}
coef(fit)
```

Alternatively, we can use the `tidySEM` function `table_results()`:

```{r}
table_results(fit)
```
